---
title: "Proteomics Data Analysis - Melanoma"
author: "Ellis Patrick"
date: "3 July, 2021"
output:
  html_document:
    number_sections: no
    theme:  spacelab
    toc: true
    toc_depth: 3
    toc_float: true
    collapsed: true
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, error = FALSE, message = FALSE, cache = TRUE)
```


## Read in data

```{r}
library(readxl)
library(limma)
library(tidyverse)
library(Biobase)
library(Glimma)
library(plotly)
library(e1071)
library(BiocParallel)
library(plotly)
library(ggforce)
```


- Changed "Filtered" to NA. It might be more appropriate to change these to zero.
```{r}
# Read in DIA
data <- read_xlsx("Melanoma_ProteinPivot_DirectDIA_May2021_GeneName_Final_1.xlsx") %>%
    as.data.frame()
data[data=="Filtered"] <- NA
data[,-1:-2] <- sapply(data[, -1:-2], as.numeric)
rownames(data) <- data$Gene
data <- data[, -c(1:2)] 


# Read in metadata

dataMeta <- read_xlsx("MetaData.xlsx", col_names = FALSE, sheet = 2) %>% as.data.frame()
colnames(dataMeta) <- c("Sample", "Condition")


```

## Normalisation check

- Data appears reasonably well normalised within datasets.
- I won't do any normalisation, but this is something we could/should revisit.

```{r}

plot(data[,c("Primary_01")], data[,c("Primary_21")], log = 'xy', main = "Primary_01 vs Primary_02")
abline(0,1)
abline(h = 1000, v = 1000, lty = 3)

```

## Should I impute?

- There is a reasonably strong relationship between the number of zeros and protein abundance.
- I will replace "Filtered" with a suitable number (1000).


```{r}

plot(rowMeans(data,na.rm = TRUE), rowSums(is.na(data)), log = 'x')

```

## Explore with PCA



### DIA data

- There appears to be some separation between the groupls in PC2.
- There appears to be a relationship with the number of "Filtered" values in PC1. This implies we should come back and try harder with out imputation/normalisation.

```{r}
dat <- data
dat[is.na(dat)] <- 1000
dat[dat<1000] <- 1000
dat <- dat[apply(dat,1,sd)!=0,]
pcaDIA <- prcomp(t(log2(dat)), scale = TRUE)

df <- data.frame(pcaDIA$x, Sample = rownames(pcaDIA$x), numZerosInSample = colSums(dat==1000)) %>%
  left_join(dataMeta, "Sample" )

ggplot(df, aes(PC1, PC2, colour = numZerosInSample)) + geom_point() + theme_classic()

ggplot(df, aes(PC1, PC2, colour = Condition)) + geom_point() + theme_classic()

```

## Differential abundance

- There are lots of DE proteins between all the conditions. Look at the Glimma html files in the Glimma folder for more info.
- Most of the proteins are different between Regional vs Primary are the same as Distant vs Primary.

```{r}
dat <- data
dat[is.na(dat)] <- 1000
dat[dat<1000] <- 1000
dat <- log2(dat)



design <- model.matrix(~Condition-1, dataMeta[-grep("Pooled",dataMeta$Condition),])
rownames(design) <- dataMeta[-grep("Pooled",dataMeta$Condition),]$Sample
colnames(design) <- gsub(" ", "", colnames(design))
colnames(design) <- gsub("Condition", "", colnames(design))

fit <- lmFit(dat[, rownames(design)], design)
contrast <- makeContrasts("DistantMetastaticMelanoma - PrimaryMelanoma", "RegionalMetastaticMelanoma - PrimaryMelanoma", "DistantMetastaticMelanoma - RegionalMetastaticMelanoma", "DistantMetastaticMelanoma - (PrimaryMelanoma + RegionalMetastaticMelanoma)/2", "PrimaryMelanoma - (DistantMetastaticMelanoma + RegionalMetastaticMelanoma)/2", "RegionalMetastaticMelanoma - (PrimaryMelanoma + DistantMetastaticMelanoma)/2", levels = colnames(design))
fit <- contrasts.fit(fit, contrast)
fit <- eBayes(fit, trend = TRUE, robust = TRUE)
apply(decideTests(fit), 2, table)


for(i in colnames(coef(fit))){

i2 <- make.names(i)
i2 <-  unlist(lapply(strsplit(i2,"\\.\\.\\."), function(x)paste(paste(x[1],x[2],sep = ".Vs."),x[3], sep = ".And.")))
i2 <- gsub("\\.And\\.NA", "", i2)
i2 <- gsub("\\.\\.2", "", i2)
i2 <- gsub("\\.\\.", "_", i2)
i2 <- gsub("\\.", "_", i2)

  topDIA <- topTable(fit,coef = i, n = 1000000, p.value = 1, adjust.method = "fdr")
dt <- decideTests(fit,p.value=0.05)
write.csv(topDIA, paste0("top",i2,".csv"))


glMDPlot(fit, coef=i, status=dt, main=i,
side.main="Gene", counts=dat, groups=dataMeta$Condition, launch=FALSE, xlab = "Average expression", samples = colnames(dat),html = i2)
}

```

### Genes that are changing

```{r}
df <- data.frame(signif(fit$t,2))
colnames(df) <- unlist(lapply(strsplit(colnames(df),"\\.\\.\\."), function(x)paste(paste(x[1],x[2],sep = ".Vs."),x[3], sep = ".And.")))
colnames(df) <- gsub("\\.And\\.NA", "", colnames(df))
colnames(df) <- gsub("\\.\\.2", "", colnames(df))
colnames(df) <- gsub("\\.\\.", "_", colnames(df))
colnames(df) <- gsub("\\.", "_", colnames(df))
df <- data.frame(df, gene = rownames(fit$t))
p1 <- ggplot(df, aes(x = DistantMetastaticMelanoma_Vs_PrimaryMelanoma, y = RegionalMetastaticMelanoma_Vs_PrimaryMelanoma, colour = PrimaryMelanoma_Vs_DistantMetastaticMelanoma_And_RegionalMetastaticMelanoma, label = gene)) + geom_point()  +  theme_classic() + scale_colour_continuous(type = "viridis") + labs(colour = "PrimaryMelanoma\nVs\nDistantMetastaticMelanoma\nAnd\nRegionalMetastaticMelanoma")
ggplotly(p1)
```



```{r, fig.height=10, fig.width = 10}
library(directPA)
data(Pathways)

```

### Pathways that are changing


```{r}
Tc <- fit$t[, 1:2]
annotation <- Pathways.reactome
minSize = 5

Tc.zscores <- apply(Tc, 2, function(x) {
        qnorm(rank(x)/(nrow(Tc) + 1))
})
    
DE = lapply(annotation, function(x) {
    if (sum(rownames(Tc.zscores) %in% x) >= minSize) {
        X <- Tc.zscores[rownames(Tc.zscores) %in% x, ]
        n = nrow(X)
        Z1 = sum(X[, 1])/sqrt(n)
        Z2 = sum(X[, 2])/sqrt(n)
        list(Z1 = Z1, Z2 = Z2)
        }
  })

DE <- DE[which(sapply(DE, length) != 0)]
Z1 <- unlist(sapply(DE, function(x) {
        x[1]
    }))
Z2 <- unlist(sapply(DE, function(x) {
        x[2]
    }))



circleFun <- function(center = c(0,0),diameter = 1, npoints = 100){
    r = diameter / 2
    tt <- seq(0,2*pi,length.out = npoints)
    xx <- center[1] + r * cos(tt)
    yy <- center[2] + r * sin(tt)
    return(data.frame(x = xx, y = yy))
}

circ2 <- data.frame(circleFun(diameter = 2*2), Pathway = NA)
circ3 <- data.frame(circleFun(diameter = 3*2), Pathway = NA)
circ4 <- data.frame(circleFun(diameter = 4*2), Pathway = NA)

df <- data.frame(Z1 = signif(Z1,3), Z2 = signif(Z2, 3), Pathway = gsub( "\\.Z1","", names(Z1)))

p1 <- ggplot(df, aes(x = Z1, y = Z2, labels = Pathway)) + geom_point()  + theme_bw() + geom_path(data = circ2, aes(x = x, y = y)) + geom_path(data = circ3, aes(x = x, y = y)) + geom_path(data = circ4, aes(x = x, y = y)) + coord_fixed() + labs(x=colnames(Tc)[1], y = colnames(Tc)[2])
ggplotly(p1)


```



## Classification

- All groups separate quite well. Reasonably high accuracy.
- These accuracies might be improved by better normalisation and/or optimising classification approach.
- Here I have just used SVM, we should try RandomForest and logistic regression.

```{r}



CV <- function(X,Y,k,seed){

set.seed(seed)
n <- length(Y) # The number of observations
nFolds <- 10
nRep <- 100

# Create a vector with the same amount of fold labels
fold <- rep(1:nFolds, 100) 
fold <- fold[1:n]

# Create matrix to store repeated CV error rates
errorRepCV <- matrix(NA, nRep, n)

for(j in 1:nRep){
# Reorder these to avoid systematic ordering bias
    foldSample <- sample(fold, n, replace = FALSE)

    predCVLoop <- factor(rep(NA, n), levels = names(table(Y)))

    for(i in 1:nFolds){
        # Split into training and test set    
        dataTest <- X[foldSample == i, ]
        dataTrain <- X[foldSample != i, ]
        outcomeTrain <- Y[foldSample != i]
        
        design <- model.matrix(~outcomeTrain)
        fit <- lmFit(t(dataTrain), design)
        efit <- eBayes(fit, trend = TRUE)
        
        topGenes <- names(sort(efit$p.value[,2]))[seq_len(k)]
        
        dataTrain <- data.frame(outcomeTrain, dataTrain[,topGenes])

        # Fit model on dataTrain
        
        fit <- svm(outcomeTrain~., dataTrain)

        # Predict on dataTest
        pred <- predict(fit, dataTest)
        

        # Save test predictions
        predCVLoop[foldSample == i] <- pred
    }
    errorRepCV[j,] <- predCVLoop == Y
}

(rowMeans(errorRepCV[,Y==levels( predCVLoop)[1]]) + rowMeans(errorRepCV[,Y==levels( predCVLoop)[2]]))/2

}




dat <- data
dat[is.na(dat)] <- 1000
dat[dat<1000] <- 1000
dat <- log2(dat)
Genes <- dat
Y <- dataMeta$Condition
Genes <- as.matrix(Genes)
rownames(Genes) <- gsub(";","",rownames(Genes))
rownames(Genes) <- gsub("-","",rownames(Genes))
X <- t(Genes)

#mean(errorRepCV)

```


### Distant vs Primary

```{r}

use <- Y %in% c("Distant Metastatic Melanoma", "Primary Melanoma")
# errorRepCV <- CV(X[use,], factor(Y[use]), 100, 51773)
# boxplot(errorRepCV)

k <- as.list(c(4,10,20,50,100, 200))
names(k) <- c(4,10,20,50,100, 200)
CVn <- function(n)CV(X[use,], factor(Y[use]), n, 51773)

errorRepCVDistant.K <- lapply(k, CVn)

boxplot(errorRepCVDistant.K, xlab = "Number of proteins", main = "Distant vs Primary")
```


### Regional vs Primary

```{r}

use <- Y %in% c("Regional Metastatic Melanoma", "Primary Melanoma")
# errorRepCV <- CV(X[use,], factor(Y[use]), 100, 51773)
# boxplot(errorRepCV)

k <- as.list(c(4,10,20,50,100, 200))
names(k) <- c(4,10,20,50,100, 200)
CVn <- function(n)CV(X[use,], factor(Y[use]), n, 51773)

errorRepCVDistant.K <- lapply(k, CVn)

boxplot(errorRepCVDistant.K, xlab = "Number of proteins", main = "Regional vs Primary")
```




### Regional vs Distant

```{r}

use <- Y %in% c("Regional Metastatic Melanoma", "Distant Metastatic Melanoma")
# errorRepCV <- CV(X[use,], factor(Y[use]), 100, 51773)
# boxplot(errorRepCV)

k <- as.list(c(4,10,20,50,100, 200))
names(k) <- c(4,10,20,50,100, 200)
CVn <- function(n)CV(X[use,], factor(Y[use]), n, 51773)


errorRepCVDistant.K <- lapply(k, CVn)

boxplot(errorRepCVDistant.K, xlab = "Number of proteins", main = "Regional vs Distant")
```

